Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2010-10-08 pkg/paper/Draft/draft-Z-W.tex
title suggestion, used for workshop
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@35 b679b2f7-6539-45d0-a605-5ddcb976cde9
updated draft and slides (with graphs) and pdf version
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@37 b679b2f7-6539-45d0-a605-5ddcb976cde9

\title{Something about Euro, EMU, and Inflation}

@@

\Abstract{
  The aim of this paper is to shed some light on the effect of the European Monetary Union (EMU)
  on changing the volatility of the inflation rates in its Member States. 
}

@@

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstraße 15\\
  6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org}
}

@@

below 2\%''. \citet{emerson} emphasizes the fact that a high inflation rate is also more variable and uncertain and thus causes more relative price variability, leading to a less efficient price mechanism (p.~22). \\

@@

The question of interest centers around the way in which a countries decision to join the EMU changed its inflation rate dynamics. 

@@

Inflation is measured as the logarithm of the monthly change in the HICP from 1.1990---3.2010. Countries included are Austria,	Belgium, the	Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	the Netherlands,	Poland,	Portugal,	Spain,	Sweden, and	the	United Kingdom.  The data are obtained from the OECD Statistics.\footnote{Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity.} \\
The countries can be divided into three different groups: the EURO countries (Austria, Belgium, Estonia\footnote{although Estonia enters in 2011}, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, the Netherlands, Portugal, Spain and Slovenia)\footnote{Cypria, Malta, Andorra, Monaco, San Marino and the Vatican are left out due their minor importance.}, EU members without ERM~II (Exchange Rate Mechanism): the Czech Republic, Hungary, Poland, the United Kingdom and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU.

@@

with location ($\theta$), scale ($\sigma$) and shape ($\delta$). For b=1 the distribution simplifies to the logistic distribution, for b$<$1 it is skewed to the left and for b$>$1 it is skewed to the right. The moments are given by:

@@

%We use the Cram{\'e}r von Mises type test as given in \citet{z07}. The test statistic is given by:
%
%\begin{eqnarray}
%& & \frac{1}{n} \sum_{i=1}^n \Vert efp(\frac{i}{n}) \Vert_2^2
%\end{eqnarray}
%
%i.e., first the $L_2$ norm is used to aggregate over the components and then the mean of the resulting aggregated process is used as the test statistic. This can also be shown graphically.
%
%
%
%\subsubsection[The chi-squared test]{The $\chi^2$ test}
%
%The main problem is to determine the number of classes k (data is grouped into k classes where we then calculate the difference between observed and expected frequencies). In a continuous distribution case, there are no natural boundaries. 
%As given in \citet{kendal} the maximum likelihood estimation of the parameters is not a big problem if k is large enough. The classes are taken to cover equal ranges of the variate, which can be done easily once k is determined. This procedure was advocated by Mann and Gumberl. One rule would be to let k be equal to  $3.765(N-1)^{2/5}$ for a 5\% significance level, when using intervals with equal probability under $H_0$, as given in \citet{boero}. We use this rule with a range of [3,2*k-3] for the tests and report any pvalues below 0.05.

@@

@@

On the basis of the results, we can trace out 6 groups of countries with rather distinct behavior. Group 1 consists of Italy and Spain (both very similar) and possibly Belgium, Luxembourg and Portugal. Group 2 is an Eastern European group comprising Poland, the Czech Republic, Slovenia, Slovakia and Hungary. Group 3 is a Non-ERM~II member group including Sweden and the United Kingdom. The Netherlands, Greece and Finland build group 4 which exhibits no changes. Group 5 is composed of France, Germany, Austria and Denmark. Ireland stands on its own and is the only country which changed its inflation dynamics as an after march of the financial crises of 2008. \\
What we can observe for almost all the EURO countries is an increase in variance and mean, with the exception of Ireland. Since inflation volatility increased in all countries, we find no proof of a common claim that inflation uncertainty may increase in countries that have a smaller influence on ECB policy. \\
Lets consider a view countries in broader detail to get an idea about what happened. In Austria we see evidence of a change in the inflation rate, when the old regime ended in September 2007. The new regime is much more volatile, with an anlmost threefold increase in variance.

@@

The reason for this is a strong increase in the yearly Austrian inflation rate following a heavy increase in oil price which was further amplified by a rise of minearl taxes. However, we fail to find any evidence for an incrase in mean or volatility of the Austrian rate following the introduction of the Euro paper money. \\
Lets consider Slovenia as the first Eastern European country that introduced the Euro. In August 2003 the new regime started with a much lower mean value (only a third now of its previous value) and interestingly enough, a higher volatility.

@@

Slovenia is a good example of the benefits of an early Euro adoption. Its efforts to meet the criterias as set forth in the Maastricht treaty were successfull, actually Slovenia reached them in 2005. So from 2003 onwards inflation was lower but with a much higher variance. Regarding the dating in 2003, we have to keep in mind that this was the year where most of the financial reforms were introduced. Slovenia seems to be the only case where the introduction of the Euro was followed by an increase in yearly inflation, which faded away following the financial crises and its deflationary effects. \\
As a last example we focus on Spain.

@@

\section{Conlcusion}

\bibliography{papers}

>>>>

\title{Structural Breaks in Inflation Dynamics within the European Monetary Union}

@@

\Abstract{
  The aim of this paper is to shed some light on the effect of the European Monetary Union (EMU)
  on changing the dynamics of the inflation rates in its Member States. 
}

@@

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstraße 15\\
  6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
}

@@

below 2\%''\footnote{http://www.ecb.europa.eu/mopo/strategy/pricestab/html/index.en.html}. \citet{emerson} emphasizes the fact that a high inflation rate is also more variable and uncertain and thus causes more relative price variability, leading to a less efficient price mechanism. \\

@@

The question of interest centers around the way in which a countries decision to join the Economic and Monetary Union (EMU) changed its inflation rate dynamics. 

@@

Inflation is measured as the logarithm of the monthly change in the HICP from 1.1990---3.2010. Countries included are Austria,	Belgium, the	Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	the Netherlands,	Poland,	Portugal,	Spain,	Sweden, and	the	United Kingdom.\footnote{Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity.} The data is obtained from the OECD Statistics.\footnote{http://www.oecd-ilibrary.org/content/data/data-00285-en} \\
The countries can be divided into three different groups: the EURO countries (Austria, Belgium, Estonia\footnote{although Estonia enters in 2011}, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, the Netherlands, Portugal, Spain and Slovenia)\footnote{Cypria, Malta, Andorra, Monaco, San Marino and the Vatican are left out due their minor importance.}, EU members without ERM~II (Exchange Rate Mechanism): the Czech Republic, Hungary, Poland, the United Kingdom and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU.

@@

with location ($\theta$), scale ($\sigma$) and shape ($\delta$). For $\delta$=1 the distribution simplifies to the logistic distribution, for $\delta$$<$1 it is skewed to the left and for $\delta$$>$1 it is skewed to the right. The moments are given by:

@@

We use the Cram{\'e}r von Mises type test as given in \citet{z07}. The test statistic is given by:

%\begin{eqnarray}
%& & \frac{1}{n} \sum_{i=1}^n \Vert efp(\frac{i}{n}) \Vert_2^2
%\end{eqnarray}
%
%i.e., first the $L_2$ norm is used to aggregate over the components and then the mean of the resulting aggregated process is used as the test statistic. This can also be shown graphically.
%
%
%
%\subsubsection[The chi-squared test]{The $\chi^2$ test}
%
%The main problem is to determine the number of classes k (data is grouped into k classes where we then calculate the difference between observed and expected frequencies). In a continuous distribution case, there are no natural boundaries. 
%As given in \citet{kendal} the maximum likelihood estimation of the parameters is not a big problem if k is large enough. The classes are taken to cover equal ranges of the variate, which can be done easily once k is determined. This procedure was advocated by Mann and Gumberl. One rule would be to let k be equal to  $3.765(N-1)^{2/5}$ for a 5\% significance level, when using intervals with equal probability under $H_0$, as given in \citet{boero}. We use this rule with a range of [3,2*k-3] for the tests and report any pvalues below 0.05.

@@

% mit cecchetti Paper im Unterlagen Ordner abgleichen

@@

On the basis of the results, we can trace out 6 groups of countries with rather distinct behavior. Group 1 consists of Italy and Spain (both very similar) and possibly Belgium, Luxembourg and Portugal. Group 2 is an Eastern European group comprising Poland, the Czech Republic, Slovenia, Slovakia and Hungary. Group 3 is a Non-ERM~II member group including Sweden and the United Kingdom. The Netherlands, Greece and Finland build group 4 which exhibits no changes. Group 5 is composed of France, Germany, Austria and Denmark. Ireland stands on its own and is the only country which changed its inflation dynamics as an after march of the financial crises of 2008. \\
What we can observe for almost all the EURO countries is an increase in variance and mean, with the exception of Ireland. Since inflation volatility increased in all countries, we find no proof of a common claim that inflation uncertainty may increase in countries that have a smaller influence on ECB policy. \\
Lets consider a few countries in broader detail to get an idea about what happened. In Austria we see evidence of a change in the inflation rate, when the old regime ended in September 2007. The new regime is much more volatile, with an almost threefold increase in variance.

@@

The reason for this is a strong increase in the yearly Austrian inflation rate following a heavy increase in oil price which was further amplified by a rise of mineral taxes. However, we fail to find any evidence for an incrase in mean or volatility of the Austrian rate following the introduction of the Euro paper money. \\
Lets consider Slovenia as the first Eastern European country that introduced the Euro. In August 2003 the new regime started with a much lower mean value (only a third now of its previous value) and interestingly enough, a higher volatility.

@@

Slovenia is a good example of the benefits of an early Euro adoption. Its efforts to meet the criterias as set forth in the Maastricht treaty were successfull, actually Slovenia reached them in 2005. So from 2003 onwards inflation was lower but with a much higher variance. Regarding the dating in 2003, we have to keep in mind that this was the year where most of the financial reforms were introduced. Slovenia seems to be the only case where the introduction of the Euro was followed by an increase in yearly inflation, which faded away following the financial crises and its deflationary effects. \\

%As a last example we focus on Spain. \\

%Other possible explanations: 
%
%- rise in value added tax (VAT; as in Germany in 2007)
%- change in central bank policy (for example Spain changed to inflation targeting in 1995; Uk, Swe and Finland in 1992)
%- role of budget deficits

% Quelle für diese Behaupunten: http://www.oenb.at/isaweb/dyna2.do?go=initIndikatoren&definitionLaden=false&hierarchieIdSelected=108

Another possible explanation pattern could be the growth in the  money supply. A drawback is the current lack of data\footnote{hopefully to be solved}. But for some countries we can find some plausible findings. Regarding the case of Denmark, the M1 money supply started contracting in June 2000 (on a monthly base). In Estonia we find the highpoint in M3 expansion ending with March 1998, with much lower M3 growth rates afterwards. The same happended to Hungary in April and May 1998. In the Polnish case we find a severe contraction of the money supply in May 2001. Similar findings hold true for Germany as well. \\
In Spain we see a stronger increase in M1 in May and June 1996 (nothing for the other break, data scarceity). In Portugal we find a modest increase in year on year M1 money supply growth starting in July 1992. We find nothing peculiar for Slovakia and Hungary. In Estonia the M1 money supply starts to contract severely on a yearly base following March 1998. In Slovenia we find a contraction of the Money Supply in July 2003, with the whole year 2003 witnessing a sharp decrease in the year on year growth. Poland starts contracting the year on year M1 money supply after June 2000 with very low growth rates till January 2002. 
\\
[irgendwo noch inne (nach \cite{duarte}), dass es normal isch, dass die relativen Preise fluktuieren, got uf asymmetrische productivity shocks zruck]

% for denmark: see: C:\Dissertation\Daten\DSI\M1\daten - the denmark analysis, alos: Spain, Portugal, Slovakia, Hungary Estonia noch am absatz für M1

Concerning monetary unions, it is interesting to notice that Belgium and Luxembourg, that constituted a monetary union prior to the European Monetary Union, did not experience the same timing in their break towards a higher mean and a significantly higher volatility regime. \\

With regard to the alleged benefit of monetary unions, the import of credibility to former high inflation countries (like Italy and Spain), did nothing to decrease their mean or volatility, rather to the contrary. Spain and Italy did have a -- rather -- low mean low volatility regime from 1996:6 to 2000:12 but then things changed towards higher volatility. It appears as if both countries tried to decrease inflation prior to the fixing of the exchange rates and let go afterwards.  \\

Following the reasoning of a study by \cite{duarte}, high volatility of inflation rates within regions of a union may be explained a countries decision to lower the inflation differential by means of fiscal policy. This explanation might fit to Spain and Italy, both were above the the average EU inflation. \\

% Quelle: für Deutschland M3 bis 1995
%http://www.bundesbank.de/statistik/statistik_zeitreihen.php?lang=de&open=ewu&func=row&tr=TSD304
%
%Deutschland: M3 bis 1995
%dort ersichtlihc, dass Deutschland zu Bruchzeitpunkten jeweils änderungen im m3 aggregat hat

@@

\clearpage
\section{Conlcusion}

\clearpage
\bibliography{papers}

\clearpage
